Interpretable Neural Networks using EAGGA

### 2025-03-14 paper/main.tex
HPO model-free finished

\subsection{Hyperparameter Optimization (HPO)}
In contrast to model parameters, which are optimized during training, hyperparameters (HPs) are those describing the machine learning algorithm and are fixed before training.
Because they usually have a significant impact on the trained model's performance, HPs are often optimized, too.
Let $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ be a dataset consisting of $n$
tuples drawn from the data-generating distribution, i.e. $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y},\forall i=1,...,n$.
Furthermore, let $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}, (\mathcal{D},\boldsymbol\lambda)\mapsto\hat{f}$
be an algorithm that maps a given dataset $\mathcal{D}$ and HP configuration $\boldsymbol\lambda\in\boldsymbol\Lambda$ to a model $\hat{f}$.
Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed HP configuration a chosen loss function with $L$.
The goal of model training is, for fixed $\boldsymbol\lambda$, to have $\mathcal{I}_{\boldsymbol\lambda}$ find model $\hat{f}$ minimising the expected generalisation error
$GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
The goal of HPO, on the other hand, is to find $\boldsymbol\lambda$ minimising the expected generalisation error,
i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)$.
For this there is no analytical expression available, making HPO a blackbox optimization problem. \citep[pp. 2f]{10.1145/3610536}

@@

% evolutionary algorithms
Another popular class of model-free optimizers are evolutionary algorithms (EAs), which are conceptually simple and can handle even complex parameter spaces,
given appropriate implementation of operators.
EAs are iteratively evolving a population of individuals (an individual is simply a hyperparameter configuration) of size $\mu$, where in each iteration (generation),
$\lambda$\footnote{We distinguish between $\lambda\in\mathbb{N}$ when referring to the offspring size of EAs and $\boldsymbol\lambda\in\boldsymbol\Lambda$ when
referring to a hyperparameter configuration.} offspring are generated via three operators \citep[pp. 10-TODO]{genetic_algos}.\\
\textbf{Reproduction} samples an individual from the population to reproduce with probability proportional to its fitnes. For HPO, fitness usually refers to the
performance metric the model resulting from the individual's HP configuration is evaluated on.\\
\textbf{Crossover} selects two ``parents'' from the pool of reproducing individuals. An index $k$ is sampled at random and all HP values from
index $k$ on are swapped, yielding two new ``children'' individuals.\\
\textbf{Mutation} randomly changes single HP values of an individual chosen for reproduction, e.g. by adding Gaussian noise to real values or flipping bits
on binary values.\\
At the end of each generation, the EA keeps the best $\mu$ individuals, either only from the offspring (``$(\mu,\lambda)$-selection'') or more commonly from
population and offspring (``$(\mu+\lambda)$-selection''), which is also called an ``elitist'' strategy, as it's guaranteed to keep the best individual.
\citep[chap. 1.3]{feurer_hyperparameter_2019}

%- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
%- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
%  * \textbf{reproduction}: pick an individual to reproduce with prob proportional to its fitness
%  * \textbf{crossover}
%    - select 2 ``parents'' from pool of reproducing individuals
%    - for a position $k$ in their HP configs, with certain prob, swap their values after position $k$, yielding 2 new children
    % TODO: use example from book page 12
%  * \textbf{mutation}: with certain prob, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
%- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
%  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
%- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
- prominent examples
  * CMA-ES
    - ``Covariance Matrix Adapation Evolution Strategy''
    - offspring generation exclusively via multivariate normal \citep[p. 8]{hansen2023cmaevolutionstrategytutorial}
      * mean = weighted average of previous generation
      * covariance = weighted covariance of previous generation, with weighting as for mean
    - weighing scheme done to sample in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 11]{hansen2023cmaevolutionstrategytutorial}
  * differential evolution \citep[-]{differential_evolution}
    - init population randomly so that entire param space is covered
      * $D$-dim vectors $\boldsymbol{x}_{i,G}$ with $i=1,2,...,n$ for gen $G$
    - mutation
      * for each $\boldsymbol{x}_{i,G}$, generate \textbf{mutant vector} $\boldsymbol{v}_{i,G+1}=\boldsymbol{x}_{r_1,G}+F\cdot(\boldsymbol{x}_{r_2,G}-\boldsymbol{x}_{r_3,G})$
      * $r_1,r_2,r_3\in\{1,2,...,n\}$ mutually different random idx + different from $i$
      * constant $F\in[0,2]$
    - crossover
      * \textbf{trial vector} $\boldsymbol{u}_{i,G+1}=(u_{i,G+1}^{(1)},u_{i,G+1}^{(2)},...,u_{i,G+1}^{(D)})$
        - chose random index $R\in\{1,2,...,D\}$
        - sample $p\sim U[0,1]$
        - define crossover constant $\text{CR}\in[0,1]$
        - $u_{i,G+1} = \begin{dcases}
          v_{i,G+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
          x_{i,G}^{(j)} & \text{else}
         \end{dcases}$
    - selection: greedy, compare fitness of $\boldsymbol{u}_{i,G+1}$ with $\boldsymbol{x}_{i,G}$, pick better

>>>>

\subsection{Hyperparameter Optimization (HPO)}
In contrast to model parameters, which are optimized during training, hyperparameters (HPs) are those describing the machine learning algorithm and are fixed before training.
Because they usually have a significant impact on the trained model's performance, HPs are often optimized, too.
Let $\mathcal{D}\subseteq\mathcal{X}\times\mathcal{Y}$ be a dataset consisting of $n$
tuples drawn from the data-generating distribution, i.e. $(\boldsymbol{x}^{(i)}, y^{(i)})\stackrel{i.i.d.}{\sim}\mathbb{P}_{\boldsymbol{x}y},\forall i=1,...,n$.
Furthermore, let $\mathcal{I}:(\mathbb{D}\times\boldsymbol\Lambda)\rightarrow\mathcal{H}, (\mathcal{D},\boldsymbol\lambda)\mapsto\hat{f}$
be an algorithm that maps a given dataset $\mathcal{D}$ and HP configuration $\boldsymbol\lambda\in\boldsymbol\Lambda$ to a model $\hat{f}$.
Denote with $\mathcal{I}_{\boldsymbol\lambda}$ an algorithm with fixed HP configuration a chosen loss function with $L$.
The goal of model training is, for fixed $\boldsymbol\lambda$, to have $\mathcal{I}_{\boldsymbol\lambda}$ find model $\hat{f}$ minimising the expected generalisation error
$GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)=\mathbb{E}_{(\boldsymbol{x},y)\sim\mathbb{P}_{\boldsymbol{x}y}}[L(y,\mathcal{I}_{\boldsymbol\lambda}(\mathcal{D})(\boldsymbol{x}))]$
The goal of HPO, on the other hand, is to find $\boldsymbol\lambda$ minimising the expected generalisation error,
i.e. $\argmin_{\boldsymbol\lambda\in\boldsymbol\Lambda} GE(\mathcal{I}_{\boldsymbol\lambda},\mathcal{D},L)$.
For this, there is no analytical expression available, making HPO a blackbox optimization problem. \citep[pp. 2f]{10.1145/3610536}

@@

% evolutionary algorithms
Another popular class of model-free optimizers are evolutionary algorithms (EAs), which are conceptually simple and can handle even complex parameter spaces,
given appropriate implementation of operators.
EAs are iteratively evolving a population of individuals (an individual is simply a hyperparameter configuration) of size $\mu$, where in each iteration (generation),
$\lambda$\footnote{We distinguish between $\lambda\in\mathbb{N}$ when referring to the offspring size of EAs and $\boldsymbol\lambda\in\boldsymbol\Lambda$ when
referring to a hyperparameter configuration.} offspring are generated via three operators \citep[pp. 10-14]{genetic_algos}.\\
\textbf{Reproduction} samples an individual from the population to reproduce with probability proportional to its fitnes. For HPO, fitness usually refers to the
performance metric the model resulting from the individual's HP configuration is evaluated on.\\
\textbf{Crossover} selects two ``parents'' from the pool of reproducing individuals. An index $k$ is sampled at random and all HP values from
index $k$ on are swapped, yielding two new ``children'' individuals.\\
\textbf{Mutation} randomly changes single HP values of an individual chosen for reproduction, e.g. by adding Gaussian noise to real values or flipping bits
on binary values.\\
At the end of each generation, the EA keeps the best $\mu$ individuals, either only from the offspring (``$(\mu,\lambda)$-selection'') or more commonly from
population and offspring (``$(\mu+\lambda)$-selection''), which is also called an ``elitist'' strategy, as it's guaranteed to keep the best individual.
\citep[chap. 1.3]{feurer_hyperparameter_2019}

One of the most popular EA implementations is the ``Covariance Matrix Adaptation Evolution Strategy'' (CMA-ES).
It employs a multivariate normal distribution to generate offspring, for which the mean is a weighted average of the previous generation's individuals
and the covariance matrix similarly is the weighted covariance of the previous generation.
The weighing scheme for both is done in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 8-11]{hansen2023cmaevolutionstrategytutorial}.

Another popular EA approach is Differential Evolution. Its populaton is randomly initialised so the entire search space is covered.
In each generation $g$, for each $\boldsymbol\lambda_{i,g}$ with $i=1,...,n$ it creates ``mutant vectors''
$\boldsymbol\nu_{i,g+1}=\boldsymbol\lambda_{r_1,g}+F\cdot(\boldsymbol\lambda_{r_2,g}-\boldsymbol\lambda_{r_3,g})$,
where $r_1,r_2,r_3\in\{1,...,n\}$ are mutually different random indices that also different from $i$ and $F\in[0,2]$ is constant.
For crossover, it then generates a ``trial vector'' $\boldsymbol\upsilon_{i,g+1}=(\upsilon_{i,g+1}^{(1)},\upsilon_{i,g+1}^{(2)},...,\upsilon_{i,g+1}^{(N)})$ with
\begin{equation}
  \upsilon_{i,g+1}^{(j)} = \begin{dcases}
    \nu_{i,g+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
    \lambda_{i,g}^{(j)} & \text{else}
  \end{dcases}
\end{equation}
for a random index $R\in\{1,2,...,D\}$, sampled $p\sim U[0,1]$, and crossover constant $\text{CR}\in[0,1]$.
Selection is done by picking the better of $\boldsymbol\upsilon_{i,g+1}$ and $\boldsymbol\lambda_{i,g}$ with respect to fitness. \citep[p. 343]{differential_evolution}
